Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2021-06-23 paper/paper.tex
Nothing much but a start to the model description

The retinotectal projection has proved to be a deep mine of information for
the study of how the cells of the central nervous system are accurately
connected together into functional networks. This projection connects the
light-gathering cells in the retina to movement-related cells in the optic
tectum (known as the superior colliculus in mammals). Light sources
originating close to each other in the environment tend to activate retinal
cells situated close together in the eye so that an image of the environment
is formed on the retinal surface. It has been discovered that the topography
of the retina is preserved within brain regions that process this information
such that cells which are adjacent within the retina primarily excite cells
adjacent in the tectum. This indicates that during development there must
exist a mechanism which ensures the correct wiring of the axons which leave
the retina and connect to cells in the tectum.

@@

Mini-review here, which contrasts some of the modelling papers. The upshot is
that a central problem with models is not how the axons know how to get closer
to their destination, but how they know they have \emph{arrived} at their true
destination.

In a recent paper, we proposed a self-organising mechanism, based on
morphogenetic signalling gradients, which can arrange
axons growing from the thalamus into the somatosensory cortex into the well
known murine barrel cortex pattern~\cite{james_modelling_2020}. In
characterising that system, we explored the effect of various types of
noise. We found that the mechanism was robust to noise in the expression of
the signalling molecules over a wide range of amplitudes and length-scales,
and that noise in the interaction parameters, which are obtained by sampling
the signalling molecules in the source tissue (the thalamic barreloid field)
could cause topological defects. The question of noise in axon guidance has
been explored by Goodhill~\cite{goodhill_can_2016}.

@@

@@

>>>>

The retinotectal projection has proved to be a deep mine of information for
the study of how the cells of the central nervous system are accurately
connected together into functional networks. This projection connects the
light-gathering cells in the retina to movement-related cells in the optic
tectum (known as the superior colliculus in mammals). Light sources
originating close to each other in the environment tend to activate retinal
cells situated close together in the eye so that an image of the environment
is formed on the retinal surface. It has been discovered that the topography
of the retina is preserved within brain regions that process this information
such that cells which are adjacent within the retina primarily excite cells
adjacent in the tectum. This indicates that during development there must
exist a mechanism which ensures the correct arrangement of the axons which
leave the retina and connect to cells in the tectum.

@@

Mini-review here, which contrasts some of the modelling papers. The upshot is
that a central problem with models is not how the axons know how to get closer
to their destination, but how they know they have \emph{arrived} at their true
destination. Also make the point that many of the models are phenomenological
in nature and do not elucidate the mechanisms behind the arrangement.

In a recent paper, we proposed a self-organising mechanism, based on
morphogenetic signalling gradients, which can arrange
axons growing from the thalamus to the somatosensory cortex into the well
known murine barrel cortex pattern~\cite{james_modelling_2020}. In
characterising that system, we explored the effect of various types of
noise. We found that the mechanism was robust to noise in the expression of
the signalling molecules over a wide range of amplitudes and length-scales,
and that noise in the interaction parameters, which are obtained by sampling
the signalling molecules in the source tissue (the thalamic barreloid field)
could cause topological defects. The question of noise in axon guidance has
been explored by Goodhill~\cite{goodhill_can_2016}.

@@

Another approach: We present a model of gradient-following based on graded
interactions between the growth cones of retinal ganglion cell axons and the
tectal surface. We show that a combination of exponentially graded expression
in the retina and logarithmically graded expression on the tectum permits a
purely local model to reproduce the majority of reported phenomena associated
with retinotectal organisation.

@@

Our model consists of agents, representing axonal growth cones, originating
from a square retina and moving upon a square tectum. Each agent carries a set
of 4 receptors and 4 ligands which are expressed at differing levels depending
upon the soma's location on the retina. The expression levels of the receptors
and ligands form gradients across the retinal surface. Receptors are arranged
in orthogonal pairs with $r_0$ being orthogonal to $r_1$ and $r_2$, whose
gradient is opposite to $r_0$ is orthogonal to $r_3$. Ligands are also
arranged in orthogonal pairs, with...
%
Similarly, the tectum expresses the same sets of
receptors and ligands, also in orthogonal pairs of gadients, though not
necessarily with the same expression function.
